The Heavy Photon Search Test Detector

### 2014-03-10 hps-testrun-nim.tex
Simplified the trigger section and moved the SVT DAQ subsection to the DAQ section.

\subsection{Digitization, Event Building and Data Transmission}
\label{sec:svt_daq}
The SVT data acquisition (DAQ) is based on the Reconfigurable Cluster Element (RCE) and cluster 
interconnect concept developed at SLAC as generic building blocks for DAQ systems. 
The RCE is a generic computational building block, housed on a separate daughter card called 
Data Processing Module (DPM), that are realized on an ATCA front board called the Cluster On Board 
(COB), see Fig.~\ref{fig:cob}.
 \begin{figure}[]
\begin{center}
{\small
	\includegraphics[width=6cm]{figures/svt_daq_module}
	\caption{The SVT DAQ COB board with four data processing daughter cards visible on the left side.}
	\label{fig:cob}
}
\end{center}
\end{figure}
The first generation RCE used in the Test run consisted of a Virtex~5 FPGA with 1~GB of DDR3 RAM. 
A schematic overview of the system is shown in Fig.~\ref{fig:svtdaq}. 
 \begin{figure}[]
\begin{center}
{\small
	\includegraphics[width=6cm]{figures/svt_daq_module}
	\caption{Block diagram overview of the SVT DAQ.}
	\label{fig:svtdaq}
}
\end{center}
\end{figure}
The outputs of 12 SVT half-modules are digitized on the Rear-Transition-Module (RTM), a custom board 
on the back side of the ATCA crate, interfacing the HPS readout to the generic DAQ components on the 
COB. A pre-amplifier converts the APV25 differential current output to a different voltage output scaled to 
the sensitive range of a 14-bit ADC operating at the system clock of 41.667~MHz. The RTM is organized 
into four sections with each section supporting three SVT half-module hybrids (15 APV25 ASICs). The 
RTM also includes a 4-channel fiber optic module and supporting logic which is used to interface 
to the JLab trigger system supervisor. Each section of the RTM is input to a DPM which apply thresholds 
for data reduction and organizes the sample data into UDP datagrams. The DPM also hosts an I$^{2}$C 
controller used to configure and monitor the APV25 ASICs. A single ATCA crate with two COB cards 
was used, one supporting four DPMs and one supporting 3 DPMs and one DPM that is configured as 
the trigger and data transmission module. The two COB cards and their DPMs are interconnected with a 
10Gb/s switch card~\cite{Larsen:2011zb} which also hosts two 1Gb/s Ethernet interfaces to the external 
SVT DAQ PC.

The external PC supports three network interfaces; two standard 1Gb/s Ethernet and one custom low 
latency data reception card. The first is used for slow control and monitoring of the 8 
DPM modules and the second serves as the interface to the JLAB data acquisition system. The third 
custom low latency network interface is used to receive data from the ATCA crate and supports a low 
latency, reliable TTL trigger acknowledge interface to the trigger DPM. This PC hosts the SVT control 
and monitoring software as well as the JLab Read Out Controller (ROC) application described in more 
detail in Sec.~\ref{sec:daq}.

@@

\subsection{Signal readout}
\label{sec:fadc}
After a 2:1 signal splitter, $1/3$ of an amplified APD signal ($2/3$ is sent to a timing module) is fed to a 
single channel of a JLab flash ADC (FADC) board~\cite{fadc}, 
see Fig~.\ref{fig:fadc}.
\begin{figure}[]
\begin{center}
{\small
\includegraphics[width=5cm]{figures/FADC250_Photo_001.jpg}
\caption{The JLab FADC250 VXS module.}
\label{fig:fadc}
}
\end{center}
\end{figure}
The FADC boards are high speed VXS modules digitizing up to 16 APD signals 
at 250~MHz and storing samples in 8~$\mu$s deep pipelines with 12-bit resolution. 
When a trigger is received, the part 
of the pipeline from 5 samples before and 30 after the signal crossed $\approx70$~MeV threshold are 
summed and stored in a 17-bit register for readout as described in Fig.~\ref{fig:hps_trigger_data}.
This scheme significantly compresses the data output of the FADC. During offline data analysis, a 
calibrated pedestal value is subtracted to obtain the actual summed energy.
Two 20-slot VXS crates with 14 (13) FADC boards were employed in the Test run to read out the top 
(bottom) half of the ECal.
In the Test run 385 out of 442 modules (87\%) was used in offline reconstruction, 39 modules were 
disabled or not read out (no FADC channel available, no APD bias voltage or masked out due to 
excessive noise) and 18 was masked offline due to noise. 
\begin{figure}[]
\begin{center}
{\small
\includegraphics[width=7cm]{figures/fadc_datapath_diagram.png}
\caption{The two different readout modes of the FADC used in the Test run.}
\label{fig:hps_trigger_data}
}
\end{center}
\end{figure}

@@

The HPS Test apparatus DAQ handles the acquisition of data from two sub-detectors: the 
SVT and the ECal, with two DAQ architectures. The SVT is readout with Advanced 
Telecom Communications Architecture (ATCA) hardware while the ECal uses VXS based 
hardware. The trigger system receives input from the ECal and distributes a trigger signal 
to all detector sub-systems to read out a selected event.

\subsection{Trigger system}
\label{sec:trigger}

The HPS test run trigger system is designed to select \ee{} events efficiently by 
using information from the ECal. The trigger looks for time coincidences of clusters in the top and bottom 
half of the ECal. The trigger system can be broken down into three sections as shown in 
Fig.~\ref{fig:hps_trigger_cal}:
 \begin{figure}[b]
\begin{center}
{\small
 \includegraphics[width=8cm]{figures/hps_trigger_cal}
\caption{Block diagram of the ECAL trigger system consisting of the FADC that samples and digitizes 
signals for each detector channel and sends them for cluster finding in the CTP. The CTP clusters are 
sent to the SSP where the final trigger decision is taken based on pairs of clusters in both halves of the 
ECal. The decision is sent to the Trigger Supervisor (TS) that generates the necessary signals to readout 
the sub-detectors.}
 \label{fig:hps_trigger_cal}
}
\end{center}
 \end{figure}
 \begin{itemize}
 \item FADC (pulse finding): samples and digitizes the signal pulses from each detector channel. 
Sends the measured pulse energy and arrival time to the Crate Trigger Processor (CTP),
\item CTP (cluster finding): groups FADC pulses from each half of the ECal into clusters. 
The cluster energy and arrival time are sent to the Sub-System Processor (SSP),
 \item SSP (cluster pair finding): searches for time coincidences of pairs of clusters from 
the top and bottom half of the ECal and applies topological selections.
 \end{itemize}

The time coincidence window of pairs of clusters in the top and bottom half of the ECal are 
programmable with 4~ns resolution.  As described above, the first stage of the trigger logic is 
incorporated into the FPGA's on the FADC boards. The trigger  
integrates using a "time-over-threshold" window as explained in Fig. \ref{fig:hps_trigger_data}.
For triggering purposes, every channel has: 
\begin{itemize}
% \item number of samples integrated before the threshold crossing (NSB),
 %\item number of samples integrated after the  threshold crossing (NSA),
 \item a trigger threshold, measured in ADC counts, 
 \item an offset, 
 \item a conversion factor to convert ADC counts into energy in MeV, from $50$ to $\sim1500$~MeV, with 5-bit accuracy 
 \item an energy threshold discriminator to provide aminimum energy cutoff
 \end{itemize}
Note that the threshold for the trigger path can be set independently from the readout threshold.
The offset is subtracted from ADC samples before they are integrated.  
The values reported to the CTP are the 5-bit pulse energy and the 3-bit time at which the 
pulse crossed the threshold. Data from every channel is sent to the CTP every 32~ns. With the available 
3-bit time information, the CTP looks for time coincidences of crystal signals within an 8~ns window (the 
limit is 4~ns).
The cluster finding algorithm is fast and makes use of the parallel processing capability of the FPGA's by 
simultaneously searching for 125 clusters, up to 3x3 in size, across the calorimeter crystal array, and it 
performs the following tasks:
\begin{itemize}
\item Adds together the energy for every 3x3 square of channels in the ECal
\item Hits are added together if their leading edges occurred within a programmable number of 4~ns 
clock cycles  (HPS test run used 8~ns coincidence time interval).
\item If the 3x3 energy sum is larger than the programmable cluster energy threshold and the sum is 
greater than any neighboring 3x3 windows, the CTP reports the cluster parameters to the SSP. 
\end{itemize}
The CTP evaluates all the hits in its half of the calorimeter every 4~ns. A programmable time window is 
used to allow hits that are slightly out of time with each other to be considered as part of a cluster sum. 
This is done by reporting hits when they occur and then reporting them again for the next $N$ number of 
4~ns clock cycles, where $N \in [0,7]$. This is useful to deal with skew and jitter that develop from the 
detector, cabling, and electronics. As described above, the CTP only selects the 3x3 window with the 
highest energy sum of its neighbors. This filtering is applied to deal with overlapping clusters and cases 
where the cluster is larger than a 3x3 window.

The final trigger decision is made by CTPs and the SSP is passed to the Trigger Supervisor (TS). The TS 
generates all necessary signals and controls the entire DAQ system readout through the Trigger 
Interface (TI) units. The TI units are installed in every crate that participate in the readout process.

The trigger system is free-running and driven by the 250~MHz global clock and has essentially zero 
dead time at the occupancies expected for HPS. The Trigger Supervisor can apply dead time if 
necessary, for example on a `busy' or `full' condition from the front-end electronics. The system is 
designed to handle trigger rates above 50~kHz and has a latency set to $\approx 3~\mu$s to match that 
required by the SVT APV25 chip.

During the test run, for the most part the trigger system required only a single cluster in either the top or 
bottom Ecal module. It was tested to work up to 20 kHz trigger rates.

\subsection{Data Acquisition and Online Computing}
\label{sec:daq}
For the ECal, every VXS crate contains a Readout Controller (ROC)
that collects digitized information, processes it, and sends it on to the Event Builder (EB). The ROC is a 
single blade Intel-based CPU module running DAQ software under CentOS Linux OS. For the SVT 
ATCA system, the ROC application runs on an embedded processor situated on the ATCA main board. 
The EB assembles information from the SVT and ECal ROCs into a single event which is passed to the 
Event Recorder (ER) that writes it to a RAID5-based data storage system capable of handling up to 
100~MB/s. The EB and other critical components run on multicore Intel-based multi-CPU servers. The 
DAQ network system is a network router providing high-speed connections between the DAQ 
components and the JLab computing facility. The SVT ROC, which must handle large data volumes, has 
a 10~Gbit/s link to the network router, while a 1~Gbit/s link is adequate for the ECal. A 10~Gbit/s uplink 
to the JLab computing facility is used for long-term storage.

The SVT DAQ is described in more detail in Sec.~\ref{sec:svt_daq}.

>>>>

@@

\subsection{Signal readout}
\label{sec:fadc}
After a 2:1 signal splitter, $1/3$ of an amplified APD signal ($2/3$ is sent to a timing module) is fed to a 
single channel of a JLab flash ADC (FADC) board~\cite{fadc}, 
see Fig~.\ref{fig:fadc}.
\begin{figure}[]
\begin{center}
{\small
\includegraphics[width=5cm]{figures/FADC250_Photo_001.jpg}
\caption{The JLab FADC250 VXS module.}
\label{fig:fadc}
}
\end{center}
\end{figure}
The FADC boards are high speed VXS modules digitizing up to 16 APD signals 
at 250~MHz and storing samples in 8~$\mu$s deep pipelines with 12-bit resolution. 
When a trigger is received, the part 
of the pipeline from 5 samples before and 30 after the signal crossed a programmable 
threshold (for the Test run this was set to $\approx70$~MeV) are 
summed and stored in a 17-bit register for readout as described in Fig.~\ref{fig:hps_trigger_data}.
This scheme significantly compresses the data output of the FADC. During offline data analysis, a 
calibrated pedestal value is subtracted to obtain the actual summed energy.
Two 20-slot VXS crates with 14 (13) FADC boards were employed in the Test run to read out the top 
(bottom) half of the ECal.
In the Test run 385 out of 442 modules (87\%) was used in offline reconstruction, 39 modules were 
disabled or not read out (no FADC channel available, no APD bias voltage or masked out due to 
excessive noise) and 18 was masked offline due to noise. 
\begin{figure}[]
\begin{center}
{\small
\includegraphics[width=8cm]{figures/fadc_datapath_diagram.png}
\caption{The two different readout modes of the FADC used in the Test run.}
\label{fig:hps_trigger_data}
}
\end{center}
\end{figure}

@@

The DAQ system handles acquisition of data from the ECal and SVT sub-detectors  with 
two DAQ architectures. The SVT DAQ is based on Advanced Telecom Communications Architecture 
(ATCA) hardware while the ECal uses VXS based hardware. Data from the sub-detectors are only 
readout when a trigger signal from the trigger system is received formed on input from the ECal. 
\subsection{Trigger system}
\label{sec:trigger}

The trigger system is designed to select time coincidences of electromagnetic clusters in the 
top and bottom halves of the ECal. Figure~\ref{fig:hps_trigger_cal} shows a schematic overview 
of each stage of the system. 
 \begin{figure}[b]
\begin{center}
{\small
 \includegraphics[width=8cm]{figures/hps_trigger_cal}
\caption{Block diagram of the ECAL trigger system consisting of the FADC that samples and digitizes 
signals for each detector channel and sends them for cluster finding in the CTP. The CTP clusters are 
sent to the SSP where the final trigger decision is taken based on pairs of clusters in both halves of the 
ECal. The decision is sent to the Trigger Supervisor (TS) that generates the necessary signals to readout 
the sub-detectors.}
 \label{fig:hps_trigger_cal}
}
\end{center}
 \end{figure}
Each channel on the FADC board have an 
independent data path to send a 5-bit pulse energy and 3-bit pulse arrival time information to a 
trigger processing board (CTP) in the same crate every 32~ns. Contrary to the readout path 
described in Sec.~\ref{sec:fadc}, this energy is a pedestal subtracted time-over-threshold sum with 
programmable offsets and minimum threshold discriminator for each channel. With input from all 
FADC channels, i.e. one half of the ECal, the CTP performs cluster finding and calculates cluster 
energy, shape and timing information. The 3x3 fixed-window, highly parallel, FPGA-based cluster 
algorithm simultaneously searches for up to 125 clusters with energy sum larger than 
the programmable energy threshold (in the Test run this was $\approx270$~MeV). Crystals in the 
fixed-window are included in the sum if the leading edges of the pulses occurred within a 8~ns time 
window (4~ns minimum). To take into account with clock skew and jitter throughout the system 
clustering algorithm can allow crystal hits from $N\in[0,7]$ previous or future 4~ns clock cycles to 
contribute to the cluster sum.  A separate sub-system board (SSP) receives the clusters (only the 
highest energy 3x3 window to deal with overlapping and very large clusters) from 
the top and bottom half CTP every 32~ns and searches for pairs of clusters in a 8~ns wide 
trigger window. The trigger supervisor (TS) generates all the necessary signals and controls the 
entire DAQ system readout through the trigger interface units installed in every crate that participate 
in the readout process.

The trigger system is free-running and driven by the 250~MHz global clock and has essentially zero 
dead time at the occupancies expected for HPS. The Trigger Supervisor can apply dead time if 
necessary, for example on a `busy' or `full' condition from the front-end electronics. The system is 
designed to handle trigger rates above 50~kHz and has a latency set to $\approx 3~\mu$s to match 
that required by the SVT APV25 ASIC. During the Test run, for the most part the trigger system 
required only a single cluster in either the top or bottom Ecal halves and tested to above 100~kHz 
trigger rates with low thresholds.

\subsection{SVT Data Acquisition}
%igitization, Event Building and Data Transmission}
\label{sec:svt_daq}
The SVT DAQ is based on the Reconfigurable Cluster Element (RCE) and cluster 
interconnect concept developed at SLAC as generic building blocks for DAQ systems. 
The RCE is a generic computational building block, housed on a separate daughter card called 
Data Processing Module (DPM), that are realized on an ATCA front board called the Cluster On Board 
(COB), see Fig.~\ref{fig:cob}.
 \begin{figure}[]
\begin{center}
{\small
	\includegraphics[width=6cm]{figures/svt_daq_module}
	\caption{The SVT DAQ COB board with four data processing daughter cards visible on the left side.}
	\label{fig:cob}
}
\end{center}
\end{figure}
The first generation RCE used in the Test run consisted of a Virtex~5 FPGA with 1~GB of DDR3 RAM. 
A schematic overview of the system is shown in Fig.~\ref{fig:svtdaq}. 
 \begin{figure}[]
\begin{center}
{\small
	\includegraphics[width=6cm]{figures/svt_daq_module}
	\caption{Block diagram overview of the SVT DAQ.}
	\label{fig:svtdaq}
}
\end{center}
\end{figure}
The analog outputs of up to 12 SVT half-modules (60 APV25 ASICs) are digitized on the Rear-Transition-
Module (RTM), a custom board 
on the back side of the ATCA crate, interfacing the HPS readout to the generic DAQ components on the 
COB. A pre-amplifier converts the APV25 differential current output to a different voltage output scaled to 
the sensitive range of a 14-bit ADC operating at the system clock of 41.667~MHz. The RTM is organized 
into four sections with each section supporting three SVT half-module hybrids (15 APV25 ASICs). The 
RTM also includes a 4-channel fiber optic module and supporting logic which is used to interface 
to the JLab trigger system supervisor. Each section of the RTM is input to a DPM which apply thresholds 
for data reduction and organizes the sample data into UDP datagrams. The DPM also hosts an I$^{2}$C 
controller used to configure and monitor the APV25 ASICs. A single ATCA crate with two COB cards 
was used, one supporting four DPMs and one supporting 3 DPMs and one DPM that is configured as 
the trigger and data transmission module. The two COB cards and their DPMs are interconnected with a 
10Gb/s switch card~\cite{Larsen:2011zb} which also hosts two 1Gb/s Ethernet interfaces to the external 
SVT DAQ PC.

The external PC supports three network interfaces; two standard 1Gb/s Ethernet and one custom low 
latency data reception card. The first is used for slow control and monitoring of the 8 
DPM modules and the second serves as the interface to the JLAB data acquisition system. The third 
custom low latency network interface is used to receive data from the ATCA crate and supports a low 
latency, reliable TTL trigger acknowledge interface to the trigger DPM. This PC hosts the SVT control 
and monitoring software as well as the Read Out Controller application used to interface with the 
JLab DAQ.

\subsection{General Data Acquisition and Online Computing}
\label{sec:daq}
Every crate participating in the readout process contains a Readout Controller (ROC) that 
collects digitized information, processes it, and sends it on to the event builder. For the ECal, both 
VXS crates run ROC applications in a single blade Intel-based CPU module running under CentOS 
Linux OS. For the SVT DAQ, the ROC application runs on the external PC under RHEL. 
The event builder assembles information from the ROCs into a single event which is passed to the 
event recorder that writes it to a RAID5-based data storage system capable of handling up to 
100~MB/s. The event builder and other critical components run on multicore Intel-based multi-CPU 
servers. The DAQ network system is a network router providing 10Gbit/s high-speed connection 
to the JLab computing facility for long-term storage. For the Test run, both the SVT and ECal ROC had a 
1Gbit/s link to the network router.
